---
title: "EDA of Dam or Reservoir Failure Hazard Zones from DataSF"
author: "Philippe Marchand"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The "Dam or Reservoir Failure Hazard Zone" dataset from DataSF is composed of 7 multipolygon objects, one per reservoir, representing the area at risk of flooding (based on a SFPUC model) if that reservoir fails.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(mapview)

dam_res <- read_sf("Dam or Reservoir Failure Hazard Zone_20260504")
dam_res
```
The interactive map below displays the multipolygon of each reservoir with a different color, but they don't need to be differentiated for SafeHome.

There is no numerical data to display for that layer, just the polygon areas. The dataset contains data columns *range_min* and *range_max*, but they have the same non-informative value ("0" and "Max", respectively) for all polygons.


```{r}
mapview(dam_res, zcol = "reservoir", alpha = 0.5)
```

