---
title: "EDA of 100-Year Storm Flood Risk from DataSF"
author: "Philippe Marchand"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

According to the [DataSF description](https://data.sfgov.org/Public-Safety/100-Year-Storm-Flood-Risk-Zone-July-2022-/jzu3-4yxp/about_data), the "100-Year Storm Flood Risk Zone" dataset represents areas of at least the size of an average city block, which would be likely to experience at least six inches of flooding during a 100-year storm (i.e., a storm severity level with 1% chance of occurring each year). This is based on a model of storm water runoff by SFPUC.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(raster)
library(mapview)

flood <- read_sf("100-Year Storm Flood Risk Zone (July 2022)_20260504")
flood
```

The dataset itself is a single multipolygon feature with no associated data fields. As the map below shows, the largest flooding areas correspond to the historical Mission Creek and Islais Creek. Most other areas shown at risk of flood are smaller, spanning one or a few blocks.

For SafeHome, the area covered by that multipolygon could be shown as a layer, similar to the tsunami risk area in the current live site.

```{r}
mapview(flood)
```

There is a different dataset on DataSF called "100-Year Storm Flood Elevations".

```{r, message=FALSE, warning=FALSE}
flood_elev <- read_sf("100-Year Storm Flood Elevations (December 2024)_20260504")
flood_elev
```

Compared with the previous dataset, this one includes elevation information (*contour_min* and *contour_max*, delimiting the lower and upper elevation boundary in feet; the max value is always 1 ft. above the min). However, this information mostly just reflects the elevation of the ground, as we can see by comparing the elevation value to the average value of a San Francisco Digital Elevation Model (DEM) for those same polygons.

```{r}
# SF DEM: https://earthworks.stanford.edu/catalog/stanford-nh236hj3673
dem <- raster("DEM_30m/dem30m.tif")

dem_flood <- extract(dem, flood_elev, fun = min)
plot(dem_flood[,1], flood_elev$contourmax * 0.305)
# 0.305 converts feet to meters
abline(0,1)
```

Note that elevations are in meters for the DEM, rather than feet. The diagonal represents the identity line (1:1) on that graph.

Therefore, I would not recommend showing that information elevation, instead just display the simple layer above.